---
title: "Seattle's Cycle Sharing Analysis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: ""
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning = FALSE, message = FALSE)
# Contributions by Hannah Xiao Si Laws, Sammy Pshyk, Megan Gillis
```

#**Introduction**
This report and analysis on Seattle's bicycle sharing trends. The data includes weather reports for the area, the station locations, and trips taken by cycle riders. Future explorations could include the addition of other forms of public transportation for the Seatle area to better understand commuter travel trends. We also hope to use this report to make a good grade in the fabulous Dr. Arnholt's Stats 3530 Data Science class.


We find the orignal information [here](https://proquestionasker.github.io/projects/2016/11/10/bicyclesseattle/).


This is an exploration of bicycle-sharing data in the city of Seattle, WA (USA) from October 2014 - August 2016. I hope to eventually combine this data with other forms of ride-sharing and transportation in the city, but this will be the first step.

Time to get started!  

```{r, warning = FALSE, message = FALSE}
#loading data

# For data manipulation and tidying
library(dplyr)
library(lubridate)
library(tidyr)

# For mapping
library(ggmap)
library(mapproj)

# For data visualizations
library(ggplot2)

# For modeling and machine learning
library(caret)
library(lubridate)
```

All of the data can be downloaded from the bicycle-sharing service [“Pronto!”’s website](https://www.prontocycleshare.com/data) or from [Kaggle](https://www.kaggle.com/pronto/cycle-share-dataset). This project contains 3 data sets and I’ll import and inspect each data file independently.  These data sets are not the full dataset that is used in the original site, but that doesn't stop what can be seen from the existing data!
```{r, warning = FALSE}
station <- read.csv(file = "station.csv", header = TRUE, stringsAsFactors = FALSE)

trip <- read.csv(file = "trip.csv", header = TRUE, stringsAsFactors = FALSE)

weather <- read.csv(file = "weather.csv", header = TRUE, stringsAsFactors = FALSE)
```


# **Station Locations**
Lets take a look at where the rental stations are located in Seattle and a quick peek at where they are located! 

```{r, echo = FALSE, message = FALSE, label = "stationLocs", fig.cap = "Station Locations", fig.align='center', results='hide'}
#For show (the graphs , THE EVI-DANCE) Formating the dates
station$install_date <- mdy(station$install_date)


# How many times were new stations installed?
station %>% summarise(n_distinct(install_date))
# How many stations were installed on each date
station %>% group_by(install_date) %>% summarise(count = n()) %>% 
  arrange(install_date)



station %>% summarise(n_distinct(station_id))

station_locs <- station %>% group_by(station_id) %>% select(1:4, -2)

mymap <- get_map(location = c(lat = 47.60, lon = -122.35), maptype = "roadmap", zoom = 12)

# Plot a single point for each Station ID
ggmap(mymap) + geom_point(aes(x = long, y = lat), data = station_locs, 
                          alpha = 1, color = "darkred", size = 2)
```

So it looks like all of the stations are located near the Lower Queen Anne, Belltown, International District, Capitol Hill and University of Washington areas. Let’s take a more zoomed-in look.  Example of looking at reference \@ref(fig:stationLocs).
```{r, label = "closeUp", fig.cap= "Stations map Zoomed In", fig.align='center', warning = FALSE, message = FALSE}
closeUp <- get_map(location = "Seattle", maptype = "roadmap", zoom = 13)

# Plot a single point for each Station ID
ggmap(closeUp) + geom_point(aes(x = long, y = lat), data = station_locs, 
                            alpha = 1, color = "darkred", size = 3)
```

Look at all those stations! Its hard to believe they managed to go out of business!

##**Station Installations**
Now, I wonder how many bikes can be parked at each station?
```{r, label = "currentDockCountHist", fig.align='center', fig.cap="Stations with Bikes", echo = FALSE}
#histogram showing current dock count
ggplot(data = station, aes (x = current_dockcount)) +
  geom_bar(fill = "pink", color = "black") +
  theme_classic() +
  labs( x = "Number of Bikes Per Station", y= "Count")
```

Well that’s weird, some of the stations have a dock count of 0. I’m assuming they didn’t start that way. Let’s calculate the change in dock count from station installation to August 31, 2016 and plot it on a map.  


##Change in Number of Bike Docks Per Station
Any stations with no change in number of docks are not shown here.
```{r, label = "bikesPerStation", fig.align = 'center',fig.cap="Bike Change per Station" , warning = FALSE, message = FALSE}
dock_change <- station %>% 
               group_by(station_id) %>% 
               select(station_id, long, lat, ends_with("dockcount")) %>% 
               mutate(dock_change = current_dockcount -install_dockcount)


#attempt to recreate map plot thing

closeUp <- get_map(location = "Seattle", maptype = "roadmap", zoom = 13)

ggmap(closeUp) + 
  geom_point(aes(x = long, y = lat, size = factor(dock_change), color = factor(dock_change)), data = dock_change, alpha = 0.8) + 
  guides(color = guide_legend(title = "change"), size = guide_legend(title = "change")) +
  scale_size_manual(values = 10:1)
```

Wow! Looks like quite a few stations took away bike docks and none gained any. Perhaps those stations weren’t being used very frequently. We’ll have to look at that a bit later.

##Current Station Size
I’m going to take one quick look at the current size of each station before moving on to the next dataset. Note: I did not include any stations that were closed as of August 31, 2016 in this map.
```{r, echo = FALSE, label = "currentStationSize",fig.align='center', fig.cap="Station Size", warning = FALSE, message = FALSE}
currDockSize <- station %>% 
               group_by(station_id) %>% 
               select(station_id, long, lat, ends_with("dockcount")) %>% 
               mutate(currDockSize = current_dockcount)

closeUp <- get_map(location = "Seattle", maptype = "roadmap", zoom = 13)

ggmap(closeUp) + 
  geom_point(aes(x = long, y = lat, size = factor(currDockSize), color = factor(currDockSize)), data = currDockSize, alpha = 0.8) + 
  guides(color = guide_legend(title = "change"), size = guide_legend(title = "change")) +
  scale_size_manual(values = 10:1)
```


So it looks like the biggest stations tend to be on the outskirts of the rest. Where there are several stations in close proximity, there tend to be fewer bike docks at each station. That makes sense, logically speaking. If you go to a station and there is no bike to rent, you can easily go to another nearby, assuming there is another nearby. In areas where the stations are more secluded, it’s more important that there be bikes and open spaces readily available for users.

Alright, I’m feeling good about exploring this dataset. Time to check out the trip dataset!


#**Exploring the Trips Dataset**

Let’s start with the number of trips per day since Pronto! began opening bike stations. To do that, we need to recode our start date/times as POSIXct objects. We’ll use the `lubridate` package to visualize the number of rides per day.
```{r, label = "visualRidesPerDay", fig.align = 'center', fig.cap="Time to visualize the number of rides per day", warning = FALSE, message = FALSE}
# Make the start and stop dates into POSIXct objects
trip_2 <- trip %>% mutate(start_dt = mdy_hm(starttime), stop_dt = mdy_hm(stoptime))

trip_2 <- trip_2 %>% mutate(start_date = paste(month(start_dt), 
                                               day(start_dt), year(start_dt), sep = "/"))
trip_2$start_date <- mdy(trip_2$start_date)

trip_2 <- trip_2 %>% mutate(stop_date = paste(month(stop_dt), 
                                              day(stop_dt), year(stop_dt), sep = "/"))
trip_2$stop_date <- mdy(trip_2$stop_date)


#Adding in Days of the Week as well.
trip_2$wd <- wday(trip_2$start_date, label = TRUE)


# Recode the dates
trip_2 %>% 
  
  group_by(start_date) %>%
  
  summarize(N = n()) %>%
  
  ggplot(aes(x = start_date, y = N)) + 
  
  geom_line() + 
  
  labs(x = "Date", y = "Number of trips per day") + 
  
  theme_bw() + geom_smooth()


```
Hmm, grouping by day is a little noisy. Perhaps we should try by month?


##**Plotting trips per month (by season, minus summer)**
First, we need to create a “Year-Month” variable and then we will plot. I think I’ll plot this by month but color it by season (where December, January, and February are “winter”, March, April, and May are “spring”, June, July, August are “summer”, and September, October, November are “autumn”)
```{r, label = "monthSeason", fig.cap="Trips Per Month (By Season)", fig.align='center'}
start_date_ym <- trip_2 %>% 
  mutate(ym = paste(year(start_date), 
                    month(start_date), sep = "/"), Season = ifelse(ym %in% c("2014/10", "2014/11"), "Fall",
                                                            ifelse(ym %in% c("2014/12", "2015/1", "2015/2"),
                                                                   "Winter",
                                                            ifelse(ym %in% c("2015/3", "2015/4", "2015/5"),
                                                                   "Spring", "Summer"))))

start_date_ym %>%
  group_by(ym, Season) %>%
  summarize(N=n()) %>%
  ggplot(aes(x = ym, y = N, color = Season, fill = Season, group = Season)) + 
  geom_point() +
  geom_line(group = 1) + 
  labs( x = "Date", y = "Number of Trips(per month)")
```

Well that intuitively makes sense. The number of trips taken per month increases in the spring, reaches a maximum in the summer, declines through the fall, remains fairly stable in the winter and then repeats.

##Average Trip Duration
Great! I wonder how the average trip duration fluctuates over this time period.  
Now to plot the average trip duration (in minutes) (plus or minus standard error), with colors indicating season.
```{r, label = "minutesSeason", fig.align='center', fig.cap="Average Trip Duration"}
Trip_Duration_Month <- start_date_ym %>% 
  mutate(trip_duration_min = tripduration/60) %>% 
  group_by(ym) %>% 
  select(ym, trip_duration_min, Season) %>% 
  summarise(Avg = mean(trip_duration_min), 
            sd = sd(trip_duration_min)) %>% 
  mutate(se = sd/sqrt(n()), Season = ifelse(ym %in% c("2014/10", "2014/11"), "Fall",
                                     ifelse(ym %in% c("2014/12", "2015/1", "2015/2"), "Winter",
                                     ifelse(ym %in% c("2015/3", "2015/4", "2015/5"), "Spring", "Summer"))))

Trip_Duration_Month %>%
  ggplot(aes(x = ym, y = Avg, col = Season, fill = Season, group = Season)) + 
  geom_point()+
  geom_line(aes(group = 1)) + 
  labs(x = "Date" , y = "Duration of Average Trip (minutes)") + 
  theme_bw() + 
  geom_errorbar(aes(ymin = Avg - se, ymax = Avg + se))

```

There’s surprisingly not a huge range in trip durations here.

The little bit of variation here makes logical sense. Longer trips were being taken in the spring and summer months rather than the fall and winter. It’s also notable that the spring and summer of 2016 may have shown fewer trips than the previous year, show a slight increase in average trip length.

##Number of Trips by Day of Week
I wonder if people are using this service to commute to/from work. Let’s look at the number of trips by day of the week.

Ok, so there are definitely more trips during the week than on the weekends. I wonder if this varies by season too.

```{r, eval = TRUE, label = "dayOfWeekSeasons", fig.cap="Trips during the week", fig.align='center'}
start_date_ym %>%
  group_by(wd, Season) %>%
  summarize(N = n())%>%
  ggplot(aes(x = wd, y = N, col = Season, fill = Season, group = Season)) + 
  geom_point() + 
  geom_line() + 
  theme_bw() + 
  labs(x = "Day of the Week", y= "Number of Trips", title = "Trips Per Day")
```

So it looks like usage is relatively consistent across seasons, at least as far as the number of trips are concerned.


##Number of Trips Per Time of Day

How about time of day? Are people using these around commuting times during the week and later on weekends?

```{r, echo = FALSE, label = "timeOfDaySeasons", fig.cap="Number of Trips by Time of Day", fig.align='center', warning=FALSE, message=FALSE}

trip_2 %>%
  mutate(ym = paste(year(start_date), 
                    month(start_date), sep = "/"), Season = ifelse(ym %in% c("2014/10", "2014/11"), "Fall",
                                                            ifelse(ym %in% c("2014/12", "2015/1", "2015/2"),
                                                                   "Winter",
                                                            ifelse(ym %in% c("2015/3", "2015/4", "2015/5"),
                                                                   "Spring", "Summer")))) %>%
  group_by(Hour = hour(start_dt), Season, wd) %>%
  summarize(N = n()) %>%
 ggplot(aes(x = Hour, y= N, color = Season, group = Season)) +
  geom_point() +
  geom_line() + 
  facet_wrap(~wd) + 
  theme_bw()
  labs(x = "Hour of Day", y = "Number of Trips")
```
Wow, looks like regardless of the season, people are commuting to/from work using this service (there’s a spike between 8 and 10 AM and another between 4 and 7 PM Monday through Friday). But the weekends seem to be popular between 10 AM and 10 PM.


##Number of Trips by Member Type

I wonder if different types of members (those who have a membership vs. those that bought a 24 hour or 3 day pass) vary in the number of trips they take.

If I were to guess, I’d think the short-term passes would be ideal for tourists or people looking for a quick weekend trip, whereas members may be more likely to continue using the service year-round. Let’s check out my assumptions by plotting, once again colored by season.

```{r, label = "memberAge", echo = FALSE, eval = TRUE, fig.cap="Trips by Different Member Type", fig.align='center'}
trip_2 %>%
  mutate(ym = paste(year(start_date), 
                    month(start_date), sep = "/"), Season = ifelse(ym %in% c("2014/10", "2014/11"), "Fall",
                                                            ifelse(ym %in% c("2014/12", "2015/1", "2015/2"),
                                                                   "Winter",
                                                            ifelse(ym %in% c("2015/3", "2015/4", "2015/5"),
                                                                   "Spring", "Summer")))) %>%
  filter(usertype !="")%>%
  group_by(Season, usertype, ym) %>%
  summarize(N = n())%>%
  ggplot(aes(x = ym, y = N, color = Season, group = usertype, fill = usertype)) + 
  geom_point(aes(shape = usertype), size = 3) + 
  geom_line() + 
  theme_bw() +
  labs(x = "Date", y= "Total Number of Trips Per Month")

  
```

Surprisingly (to me, at least), different types of users seem to follow similar patterns of usage. Spring and Summer are definitely the most popular times for anyone to ride a bike in the Seattle area.

##  Trip Duration by Member Type

While it may seem that the trip duration shouldn’t vary widely by member type, a quick look at Pronto!’s pricing structure may make you reconsider that assumption. You see, while you have to purchase either an annual membership ($85/year), a 24-Hour Pass ($8) or a 3-Day Pass ($16) there is still a cap on the duration of your trip. For members, any ride under 45 minutes is free, but any ride going over 45 minutes will incur a fee of $2 for every additional 30 minutes. For short-term users, any ride under 30 minutes is free, but going over that time limit would cost you an additional $2 for the first 30 minutes and $5 for each additional 30 minutes after that!

Let’s see if these time limits cause differing behaviors in our users.

```{r, label = "TripDMemType" , eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Trip Duration by Member Type", fig.align='center'}
library(tidyverse)
start_date_ym <- mutate(start_date_ym, freelim= ifelse(usertype=="Member", 45, 30))
start_date_ym %>%
  filter(usertype !="") %>%
  group_by(tripduration, trip_id, usertype, freelim) %>%
  summarize(N=n()) %>%
  ggplot(aes(x=tripduration/60, fill=usertype), group=usertype) +
    geom_histogram(binwidth = 2, show.legend = FALSE) +
    xlim(1,60) +
    facet_grid(.~usertype) + 
    scale_fill_manual(values=c("pink", "light blue")) +
    theme_bw() +
    guides(colour=FALSE) +
    geom_vline(aes(xintercept=freelim), linetype="dashed") +
    labs(x="Trip Duration (minutes)", y="Number of Trips") 
```

Ok, so our members are pretty good about making sure that they return their bike before they incur extra charges, but the short-term pass holders frequently go over their time limit. I wonder how the cost of a trip varies for members and pass holders. Let’s try to calculate the cost of a trip.

##Additional Fees

Let’s see how much these folks are paying in additional fees!

```{r, echo = FALSE, label = "AddFee", fig.align='center', fig.cap="Additional Fees by Memeber Type", warning=FALSE, message=FALSE}
start_date_ym <- mutate(start_date_ym, tripduration_m = tripduration/60)
start_date_ym <- start_date_ym %>% mutate(cost = ifelse(usertype == "Member" & tripduration_m <= 45, 0,
                                      ifelse(usertype == "Member" & tripduration_m >45 & tripduration_m <= 75, 2,
                                      ifelse(usertype == "Member" & tripduration_m > 75, (2 + 2 * ceiling((tripduration_m - 75)/30)),
                                      ifelse(usertype == "Short-Term Pass Holder" & tripduration_m <= 30, 0,
                                      ifelse(usertype == "Short-Term Pass Holder" & tripduration_m > 30 & tripduration_m < 60, 2,
                                      ifelse(usertype == "Short-Term Pass Holder" & tripduration_m > 60,
                                             (2 + 5 * ceiling((tripduration_m - 60)/30)), NA)))))))


start_date_ym %>%
  filter(cost > 0) %>%
  ggplot(aes(x = cost, fill = usertype)) + 
  geom_histogram() + 
  facet_grid(.~usertype) + 
  scale_fill_manual(values = c("purple", "pink")) + 
  guides(fill = FALSE) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  labs(x = "Additional Fees ($)")
```

Looks like short-term pass holders (who are already paying a higher price per day of biking), are also paying lots of extra fees. This could be because they are unfamiliar with the pricing structure and don’t realize they need to return their bike to a station within 30 minutes without getting charged. It is also possible that short-term users may be tourists who don’t know their way around as easily, and thus can’t find their way to a station within the time limit.

##Member Demographics

We only seem to have age and gender information about people who have an annual Pronto! membership, so we can at least take a look at what types of people use this service.

Let’s look first at age.

```{r, label = "MemDemo", eval=TRUE, fig.cap = "Age Demographic of Members", fig.align='center',  message = FALSE, warnings = FALSE, echo = FALSE}

trip_2$usertype <- as.factor(trip_2$usertype)
trip_age <- trip_2 %>% mutate(age = year(start_dt) - birthyear)

hist(trip_age$age, main = "Member Age", xlab = "Number of Riders", 
    col = "light blue", breaks = 25)

```

My first instinct here is to say “Wow! There’s a lot of 20 and 30-somethings that use this service!” But this figure (and these data) may be a little misleading. You see, we don’t have any sort of Rider ID number, meaning we can’t take “individual activity level” into account. So we can’t tell if the tallest spike is because 5 very athletic 28-year-olds went on 4,000 trips each, or if 100 people went on 200 trips each, or if there were 20,000 28-year-olds who each only used the service once.

The same problem would arise if we looked at gender, so I’m just going to move beyond demographics.







#**Exploring the Weather Dataset**
It’s time to take a brief look at the weather dataset and do a few quick visualizations.

```{r}
# Adjusting the Date Variable
weather$Date <- mdy(weather$Date)

# Adjusting the Events Variable
weather$Events <- as.factor(weather$Events)
```

```{r}
weather$Events <- gsub("Fog , Rain|Fog-Rain", "Fog-Rain", weather$Events)
weather$Events <- gsub("Rain , Snow|Rain-Snow", "Rain-Snow", 
    weather$Events)
weather$Events <- gsub("Rain , Thunderstorm|Rain-Thunderstorm", 
    "Rain-TS", weather$Events)

weather$Events <- as.factor(weather$Events)
```

```{r}
weather$Max_Gust_Speed_MPH <- gsub("-", 0, weather$Max_Gust_Speed_MPH)

weather$Max_Gust_Speed_MPH <- as.numeric(weather$Max_Gust_Speed_MPH)
```

```{r}
weather[which(is.na(weather$Mean_Temperature_F)), 1]
```


```{r}
weather[490, "Mean_Temperature_F"] <- 50
```

```{r}
weather$Events <- gsub("^$", "Other", weather$Events)
weather$Events <- as.factor(weather$Events)
```

##Temperature
**Minimum Temperature**
```{r}
weather %>% 
  
  group_by(Date) %>%
  
  
  ggplot(aes(x = Date, y = Min_TemperatureF)) + 
  
  geom_line() + 
  
  labs(x = "Date", y = "Minimum Temperature (F)") + 
  
  theme_bw() 


```

**Mean Temperature**
```{r}
weather %>% 
  
  group_by(Date) %>%
  
  ggplot(aes(x = Date, y = Mean_Temperature_F)) + 
  
  geom_line() + 
  
  labs(x = "Date", y = "Mean Temperature (F)") + 
  
  theme_bw() 
```

**Maximum Temperature**
```{r}
weather %>% 
  
  group_by(Date) %>%
  
  
  ggplot(aes(x = Date, y = Max_Temperature_F)) + 
  
  geom_line() + 
  
  labs(x = "Date", y = "Max Temperature (F)") + 
  
  theme_bw() 
```

**Events**
```{r}
weather %>%
  ggplot(aes(x = as.factor(Events)))+
  geom_bar(fill = "Green")+
  labs(x = "Events", y = "Number of Events") +
  theme_bw()
```






##Combining Trips and Weather Datasets
Good, so we can now see some parts of the weather data. Let’s combine the weather data with our trip data. Let’s try a `left join` from the `dplyr` package.
```{r}
trip_3 <- trip_2

# Change column name in trip_3 to match weather dataset
trip_3$Date <- trip_3$start_date

# Left join the trip and weather dataframes by date.
trip_weather <- left_join(trip_3, weather, by = "Date")
```


##Mean Temperature vs. Number of Trips
Ok. Now let’s see how the number of trips per day is influenced by weather (mean temperature, rounded to the nearest 5 degrees F)

```{r}
trip_weather %>%
  mutate(Mean_Temperature_F5 = 5*round(Mean_Temperature_F/5))%>%
  group_by(Mean_Temperature_F5) %>%
  summarize(N = n(), Days = n_distinct(Date)) %>%
  ggplot(aes(x = Mean_Temperature_F5, y = N)) +
    geom_line() + 
    theme_bw() + 
    labs(x = "Temperature (Rounded to Nearest 5 degrees F)", y = "Number of Trips / Number of Days at Temperature")

```

So, as expected, there are more trips when the weather is mild but not too warm (over 70F) or too cold (below 50F). However, this figure may be influenced by the overall number of days that exhibited each mean temperature. Let’s try to standardize that.

```{r}
trip_weather %>%
  mutate(Mean_Temperature_F5 = 5*round(Mean_Temperature_F/5))%>%
  group_by(Mean_Temperature_F5) %>%
  summarize(N = n(), Days = n_distinct(Date)) %>%
  mutate(Ratio = N/Days)%>%
  ggplot(aes(x = Mean_Temperature_F5, y = Ratio)) +
    geom_line() + 
    theme_bw() + 
    labs(x = "Temperature (Rounded to Nearest 5 degrees F)", y = "Number of Trips / Number of Days at Temperature")
  
```
So when we standardize our measurements, correcting for the number of days that actually reached each temperature, we see a steady increase in the number of trips until around 75F where the trend levels off. People are more likely to ride a bike when it’s warm outside.

##Precipitation vs. Number of Trips
If you’ve ever heard of Seattle, you probably hear that it rains all the time there. Let’s see if that has an impact on the number of trips taken in a day.

We’ll start with a figure standardized for number of days at a precipitation level, rounded to the nearest 0.2 inches.
```{r}
trip_weather %>%
  mutate(Precipitation_In02 = 0.2 * round(Precipitation_In / 0.2)) %>%
  group_by(Precipitation_In02) %>%
  summarize(N = n(), Days = n_distinct(Date)) %>%
  mutate(RatioPrecip = N/Days) %>%
  ggplot(aes(x = Precipitation_In02, y = RatioPrecip)) + 
  geom_line() + 
  theme_bw() + 
  labs(x = "Precipitation (Rounded to nearest 0.2 Inches)", y = "Number of Trips / Number of Days at Precipitation Level")
```


Looks like even Seattleites have a limit when it comes to riding a bike in the rain. The more it rained, the fewer trips were taken per day.








