---
title: "Seattle's Cycle Sharing Analysis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: ""
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', warning = FALSE, message = FALSE)
# Contributions by Hannah Xiao Si Laws, Sammy Pshyk, Megan Gillis
```

#**Introduction**
This report and analysis on Seattle's bicycle sharing trends. The data includes weather reports for the area, the station locations, and trips taken by cycle riders. Future explorations could include the addition of other forms of public transportation for the Seatle area to better understand commuter travel trends. We also hope to use this report to make a good grade in the fabulous Dr. Arnholt's Stats 3530 Data Science class.


We find the orignal information [here](https://proquestionasker.github.io/projects/2016/11/10/bicyclesseattle/).


This is an exploration of bicycle-sharing data in the city of Seattle, WA (USA) from October 2014 - August 2016. I hope to eventually combine this data with other forms of ride-sharing and transportation in the city, but this will be the first step.

Time to get started!  
##**Loading Necessary Packages and Importing Data**
```{r, warning = FALSE, message = FALSE}
#loading data

# For data manipulation and tidying
library(dplyr)
library(lubridate)
library(tidyr)

# For mapping
library(ggmap)
library(mapproj)

# For data visualizations
library(ggplot2)

# For modeling and machine learning
library(caret)
library(lubridate)
```

All of the data can be downloaded from the bicycle-sharing service [“Pronto!”’s website](https://www.prontocycleshare.com/data) or from [Kaggle](https://www.kaggle.com/pronto/cycle-share-dataset). This project contains 3 data sets and I’ll import and inspect each data file independently.  These data sets are not the full dataset that is used in the original site, but that doesn't stop what can be seen from the existing data!
```{r, warning = FALSE}
station <- read.csv(file = "station.csv", header = TRUE, stringsAsFactors = FALSE)

trip <- read.csv(file = "trip.csv", header = TRUE, stringsAsFactors = FALSE)

weather <- read.csv(file = "weather.csv", header = TRUE, stringsAsFactors = FALSE)
```


# **Station Locations**
Lets take a look at where the rental stations are located in Seattle and a quick peek at where they are located! 

```{r, echo = FALSE, message = FALSE, label = "stationLocs", fig.cap = "Station Locations", fig.align='center', results='hide'}
#For show (the graphs , THE EVI-DANCE) Formating the dates
station$install_date <- mdy(station$install_date)


# How many times were new stations installed?
station %>% summarise(n_distinct(install_date))
# How many stations were installed on each date
station %>% group_by(install_date) %>% summarise(count = n()) %>% 
  arrange(install_date)



station %>% summarise(n_distinct(station_id))

station_locs <- station %>% group_by(station_id) %>% select(1:4, -2)

mymap <- get_map(location = c(lat = 47.60, lon = -122.35), maptype = "roadmap", zoom = 12)

# Plot a single point for each Station ID
ggmap(mymap) + geom_point(aes(x = long, y = lat), data = station_locs, 
                          alpha = 1, color = "darkred", size = 2)
```

So it looks like all of the stations are located near the Lower Queen Anne, Belltown, International District, Capitol Hill and University of Washington areas. Let’s take a more zoomed-in look.  Example of looking at reference \@ref(fig:stationLocs).
```{r, label = "closeUp", fig.cap= "Stations map Zoomed In", fig.align='center', warning = FALSE, message = FALSE}
closeUp <- get_map(location = "Seattle", maptype = "roadmap", zoom = 13)

# Plot a single point for each Station ID
ggmap(closeUp) + geom_point(aes(x = long, y = lat), data = station_locs, 
                            alpha = 1, color = "darkred", size = 3)
```


Look at all those stations! Its hard to believe they managed to go out of business!

##**Station Installations**

```{r, label = "currentDockCountHist", fig.align='center', fig.cap="Stations with Bikes", echo = FALSE}
#histogram showing current dock count
ggplot(data = station, aes (x = current_dockcount)) +
  geom_bar(fill = "pink", color = "black") +
  theme_classic() +
  labs( x = "Number of Bikes Per Station", y= "Count")
```


Well that’s weird, some of the stations have a dock count of 0. I’m assuming they didn’t start that way. Let’s calculate the change in dock count from station installation to August 31, 2016 and plot it on a map.  


##Change in Number of Bike Docks Per Station
Any stations with no change in number of docks are not shown here.
```{r, label = "bikesPerStation", fig.align = 'center',fig.cap="Bike Change per Station" , warning = FALSE, message = FALSE}
dock_change <- station %>% 
               group_by(station_id) %>% 
               select(station_id, long, lat, ends_with("dockcount")) %>% 
               mutate(dock_change = current_dockcount -install_dockcount)


#attempt to recreate map plot thing

closeUp <- get_map(location = "Seattle", maptype = "roadmap", zoom = 13)

ggmap(closeUp) + 
  geom_point(aes(x = long, y = lat, size = factor(dock_change), color = factor(dock_change)), data = dock_change, alpha = 0.8) + 
  guides(color = guide_legend(title = "change"), size = guide_legend(title = "change")) +
  scale_size_manual(values = 10:1)
```

Wow! Looks like quite a few stations took away bike docks and none gained any. Perhaps those stations weren’t being used very frequently. We’ll have to look at that a bit later.

##Current Station Size
I’m going to take one quick look at the current size of each station before moving on to the next dataset. Note: I did not include any stations that were closed as of August 31, 2016 in this map
```{r, echo = FALSE, label = "currentStationSize",fig.align='center', fig.cap="Station Size", warning = FALSE, message = FALSE}
currDockSize <- station %>% 
               group_by(station_id) %>% 
               select(station_id, long, lat, ends_with("dockcount")) %>% 
               mutate(currDockSize = current_dockcount)

closeUp <- get_map(location = "Seattle", maptype = "roadmap", zoom = 13)

ggmap(closeUp) + 
  geom_point(aes(x = long, y = lat, size = factor(currDockSize), color = factor(currDockSize)), data = currDockSize, alpha = 0.8) + 
  guides(color = guide_legend(title = "change"), size = guide_legend(title = "change")) +
  scale_size_manual(values = 10:1)
```


So it looks like the biggest stations tend to be on the outskirts of the rest. Where there are several stations in close proximity, there tend to be fewer bike docks at each station. That makes sense, logically speaking. If you go to a station and there is no bike to rent, you can easily go to another nearby, assuming there is another nearby. In areas where the stations are more secluded, it’s more important that there be bikes and open spaces readily available for users.

Alright, I’m feeling good about exploring this dataset. Time to check out the trip dataset!


#**Exploring the Trips Dataset**

Let’s start with the number of trips per day since Pronto! began opening bike stations. To do that, we need to recode our start date/times as POSIXct objects. We’ll use the `lubridate` package to visualize the number of rides per day.
```{r, label = "visualRidesPerDay", fig.align = 'center', fig.cap="Time to visualize the number of rides per day", warning = FALSE, message = FALSE}
# Make the start and stop dates into POSIXct objects
trip_2 <- trip %>% mutate(start_dt = mdy_hm(starttime), stop_dt = mdy_hm(stoptime))

trip_2 <- trip_2 %>% mutate(start_date = paste(month(start_dt), 
                                               day(start_dt), year(start_dt), sep = "/"))
trip_2$start_date <- mdy(trip_2$start_date)

trip_2 <- trip_2 %>% mutate(stop_date = paste(month(stop_dt), 
                                              day(stop_dt), year(stop_dt), sep = "/"))
trip_2$stop_date <- mdy(trip_2$stop_date)


#Adding in Days of the Week as well.
trip_2$wd <- wday(trip_2$start_date, label = TRUE)


# Recode the dates
trip_2 %>% 
  
  group_by(start_date) %>%
  
  summarize(N = n()) %>%
  
  ggplot(aes(x = start_date, y = N)) + 
  
  geom_line() + 
  
  labs(x = "Date", y = "Number of trips per day") + 
  
  theme_bw() + geom_smooth()


```
Hmm, grouping by day is a little noisy. Perhaps we should try by month?


##**Plotting trips per month (by season, minus summer)**
First, we need to create a “Year-Month” variable and then we will plot. I think I’ll plot this by month but color it by season (where December, January, and February are “winter”, March, April, and May are “spring”, June, July, August are “summer”, and September, October, November are “autumn”)
```{r, label = "monthSeason", fig.cap="Trips Per Month (By Season)", fig.align='center'}
start_date_ym <- trip_2 %>% 
  mutate(ym = paste(year(start_date), 
                    month(start_date), sep = "/"), Season = ifelse(ym %in% c("2014/10", "2014/11"), "Fall",
                                                            ifelse(ym %in% c("2014/12", "2015/1", "2015/2"),
                                                                   "Winter",
                                                            ifelse(ym %in% c("2015/3", "2015/4", "2015/5"),
                                                                   "Spring", "Summer"))))

start_date_ym %>%
  group_by(ym, Season) %>%
  summarize(N=n()) %>%
  ggplot(aes(x = ym, y = N, color = Season, fill = Season, group = Season)) + 
  geom_point() +
  geom_line(group = 1) + 
  labs( x = "Date", y = "Number of Trips(per month)")
```



##Average Trip Duration
Great! I wonder how the average trip duration fluctuates over this time period.  
Now to plot the average trip duration (in minutes) (plus or minus standard error), with colors indicating season.
```{r, label = "minutesSeason", fig.align='center', fig.cap="Average Trip Duration"}
Trip_Duration_Month <- start_date_ym %>% 
  mutate(trip_duration_min = tripduration/60) %>% 
  group_by(ym) %>% 
  select(ym, trip_duration_min, Season) %>% 
  summarise(Avg = mean(trip_duration_min), 
            sd = sd(trip_duration_min)) %>% 
  mutate(se = sd/sqrt(n()), Season = ifelse(ym %in% c("2014/10", "2014/11"), "Fall",
                                     ifelse(ym %in% c("2014/12", "2015/1", "2015/2"), "Winter",
                                     ifelse(ym %in% c("2015/3", "2015/4", "2015/5"), "Spring", "Summer"))))

Trip_Duration_Month %>%
  ggplot(aes(x = ym, y = Avg, col = Season, fill = Season, group = Season)) + 
  geom_point()+
  geom_line(aes(group = 1)) + 
  labs(x = "Date" , y = "Duration of Average Trip (minutes)") + 
  theme_bw() + 
  geom_errorbar(aes(ymin = Avg - se, ymax = Avg + se))

```

##Number of Trips by Day of Week
```{r, eval = TRUE, label = "dayOfWeekSeasons", fig.cap="Trips during the week", fig.align='center'}
start_date_ym %>%
  group_by(wd, Season) %>%
  summarize(N = n())%>%
  ggplot(aes(x = wd, y = N, col = Season, fill = Season, group = Season)) + 
  geom_point() + 
  geom_line() + 
  theme_bw() + 
  labs(x = "Day of the Week", y= "Number of Trips", title = "Trips Per Day")
```


##Number of Trips Per Time of Day
```{r, echo = FALSE, label = "timeOfDaySeasons", fig.cap="Number of Trips by Time of Day", fig.align='center', warning=FALSE, message=FALSE}

trip_2 %>%
  mutate(ym = paste(year(start_date), 
                    month(start_date), sep = "/"), Season = ifelse(ym %in% c("2014/10", "2014/11"), "Fall",
                                                            ifelse(ym %in% c("2014/12", "2015/1", "2015/2"),
                                                                   "Winter",
                                                            ifelse(ym %in% c("2015/3", "2015/4", "2015/5"),
                                                                   "Spring", "Summer")))) %>%
  group_by(Hour = hour(start_dt), Season, wd) %>%
  summarize(N = n()) %>%
 ggplot(aes(x = Hour, y= N, color = Season, group = Season)) +
  geom_point() +
  geom_line() + 
  facet_wrap(~wd) + 
  theme_bw()
  labs(x = "Hour of Day", y = "Number of Trips")
```

##Number of Trips by Member Type
```{r, label = "memberAge", echo = FALSE, eval = TRUE, fig.cap="Trips by Different Member Type", fig.align='center'}
trip_2 %>%
  mutate(ym = paste(year(start_date), 
                    month(start_date), sep = "/"), Season = ifelse(ym %in% c("2014/10", "2014/11"), "Fall",
                                                            ifelse(ym %in% c("2014/12", "2015/1", "2015/2"),
                                                                   "Winter",
                                                            ifelse(ym %in% c("2015/3", "2015/4", "2015/5"),
                                                                   "Spring", "Summer")))) %>%
  filter(usertype !="")%>%
  group_by(Season, usertype, ym) %>%
  summarize(N = n())%>%
  ggplot(aes(x = ym, y = N, color = Season, group = usertype, fill = usertype)) + 
  geom_point(aes(shape = usertype, size = 2)) + 
  geom_line() + 
  theme_bw() + 
  labs(x = "Date", y= "Total Number of Trips Per Month")

```


##  Trip Duration by Member Type
```{r, label = "TripDMemType" , eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Trip Duration by Member Type", fig.align='center'}
library(tidyverse)
start_date_ym <- mutate(start_date_ym, freelim= ifelse(usertype=="Member", 45, 30))
start_date_ym %>%
  filter(usertype !="") %>%
  group_by(tripduration, trip_id, usertype, freelim) %>%
  summarize(N=n()) %>%
  ggplot(aes(x=tripduration/60, fill=usertype), group=usertype) +
    geom_histogram(binwidth = 2, show.legend = FALSE) +
    xlim(1,60) +
    facet_grid(.~usertype) + 
    scale_fill_manual(values=c("pink", "light blue")) +
    theme_bw() +
    guides(colour=FALSE) +
    geom_vline(aes(xintercept=freelim), linetype="dashed") +
    labs(x="Trip Duration (minutes)", y="Number of Trips") 
```


##Additional Fees
```{r, echo = FALSE, label = "AddFee", fig.align='center', fig.cap="Additional Fees by Memeber Type", warning=FALSE, message=FALSE}
start_date_ym <- mutate(start_date_ym, tripduration_m = tripduration/60)
start_date_ym <- start_date_ym %>% mutate(cost = ifelse(usertype == "Member" & tripduration_m <= 45, 0,
                                      ifelse(usertype == "Member" & tripduration_m >45 & tripduration_m <= 75, 2,
                                      ifelse(usertype == "Member" & tripduration_m > 75, (2 + 2 * ceiling((tripduration_m - 75)/30)),
                                      ifelse(usertype == "Short-Term Pass Holder" & tripduration_m <= 30, 0,
                                      ifelse(usertype == "Short-Term Pass Holder" & tripduration_m > 30 & tripduration_m < 60, 2,
                                      ifelse(usertype == "Short-Term Pass Holder" & tripduration_m > 60,
                                             (2 + 5 * ceiling((tripduration_m - 60)/30)), NA)))))))


start_date_ym %>%
  filter(cost > 0) %>%
  ggplot(aes(x = cost, fill = usertype)) + 
  geom_histogram() + 
  facet_grid(.~usertype) + 
  scale_fill_manual(values = c("purple", "pink")) + 
  guides(fill = FALSE) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  labs(x = "Additional Fees ($)")
```


##Member Demographics
```{r, label = "MemDemo", eval=TRUE, fig.cap = "Age Demographic of Members", fig.align='center',  message = FALSE, warnings = FALSE, echo = FALSE}

trip_2$usertype <- as.factor(trip_2$usertype)
trip_age <- trip_2 %>% mutate(age = year(start_dt) - birthyear)

hist(trip_age$age, main = "Member Age", xlab = "Number of Riders", 
    col = "light blue", breaks = 25)

```

```{r, eval = FALSE, echo = FALSE}
# #Members Demographic
# trip_2$usertype <- as.factor(trip_2$usertype)
# trip_age <- trip_2 %>% mutate(age = year(start_dt) - birthyear)
# 
# hist(trip_age$age, main = "Member Age", xlab = "Number of Riders", 
#      col = "#56B4E9", breaks = 25)







#Trip routes
# Create a dataframe with only station ID, latitude, and
# longitude
station_coord <- station %>% select(station_id, lat, long)

# Trim our trip dataframe to only include start & stop
# dates/times, and station ID
trip_route <- trip_2 %>% select(trip_id, starts_with("start_"), 
                                starts_with("stop_"), from_station_id, to_station_id, tripduration)

# Match by station ID
trip_route$start_lat <- station_coord[match(trip_route$from_station_id, 
                                            station_coord$station_id), "lat"]

trip_route$start_long <- station_coord[match(trip_route$from_station_id, 
                                             station_coord$station_id), "long"]

trip_route$stop_lat <- station_coord[match(trip_route$to_station_id, 
                                           station_coord$station_id), "lat"]

trip_route$stop_long <- station_coord[match(trip_route$to_station_id, 
                                            station_coord$station_id), "long"]



# Converting trip duration to minutes
trip_route$tripduration <- trip_route$tripduration/60

# Finding actual trip duration
trip_route[1, "tripduration"]


leg_1 %>% group_by(route) %>% summarise(duration = sum(minutes))

trip_route[nrow(trip_route), "tripduration"]



leg_2 %>% group_by(route) %>% summarise(duration = sum(minutes))
```





```{r, eval = FALSE, echo = FALSE}
#weather data stuff
str(weather)
# Adjusting the Date Variable
weather$Date <- mdy(weather$Date)

# Adjusting the Events Variable
weather$Events <- as.factor(weather$Events)

#Let’s combine a few of these things that seem to represent the same event.
weather$Events <- gsub("Fog , Rain|Fog-Rain", "Fog-Rain", weather$Events)
weather$Events <- gsub("Rain , Snow|Rain-Snow", "Rain-Snow", 
                       weather$Events)
weather$Events <- gsub("Rain , Thunderstorm|Rain-Thunderstorm", 
                       "Rain-TS", weather$Events)

weather$Events <- as.factor(weather$Events)


weather$Max_Gust_Speed_MPH <- gsub("-", 0, weather$Max_Gust_Speed_MPH)

weather$Max_Gust_Speed_MPH <- as.numeric(weather$Max_Gust_Speed_MPH)



weather[which(is.na(weather$Mean_Temperature_F)), 1]


weather[490, "Mean_Temperature_F"] <- "50"

weather$Events <- gsub("^$", "Other", weather$Events)
weather$Events <- as.factor(weather$Events)



#weather and trip data combo
# Make a copy of the data frame
trip_3 <- trip_2

# Change column name in trip_3 to match weather dataset
trip_3$Date <- trip_3$start_date

# Left join the trip and weather dataframes by date.
trip_weather <- left_join(trip_3, weather, by = "Date")
```











