---
title: "Seattle's Cycle Sharing Analysis"
author: "Hannah Xiao Si Laws, Sammy Pshyk, Megan Gillis"
date: "9/19/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, eval = FALSE}
"the website"
https://proquestionasker.github.io/projects/2016/11/10/bicyclesseattle/
```




```{r, echo = FALSE, message=FALSE}
#loading data

setwd("/home/students/lawshx/lawshannah.github.io/CycleShareProj")
# For data manipulation and tidying
library(dplyr)
library(lubridate)
library(tidyr)

# For mapping
library(ggmap)
library(mapproj)

# For data visualizations
library(ggplot2)

# For modeling and machine learning
library(caret)
library(lubridate)

station <- read.csv(file = "station.csv", header = TRUE, 
                    stringsAsFactors = FALSE)

trip <- read.csv(file = "trip.csv", header = TRUE, 
                 stringsAsFactors = FALSE)

weather <- read.csv(file = "weather.csv", header = TRUE, 
                    stringsAsFactors = FALSE)
```
## **Introduction**
This report and analysis on Seattle's bicycle sharing trends. The data includes weather reports for the area, the station locations, and trips taken by cycle riders. Future explorations could include the addition of other forms of public transportation for the Seatle area to better understand commuter travel trends. We also hope to use this report to make a good grade in the fabulous Dr. Arnholt's Stats 3530 Data Science class.

## **Station Locations**
Lets take a look at where the rental stations are located in Seattle!
```{r}
#For show (the graphs , THE EVI-DANCE) Formating the dates
station$install_date <- mdy(station$install_date)


# How many times were new stations installed?
station %>% summarise(n_distinct(install_date))
# How many stations were installed on each date
station %>% group_by(install_date) %>% summarise(count = n()) %>% 
    arrange(install_date)



station %>% summarise(n_distinct(station_id))

station_locs <- station %>% group_by(station_id) %>% select(1:4, 
    -2)


mymap <- get_map(location = "Seattle", maptype = "roadmap", zoom = 13)

# Plot a single point for each Station ID
ggmap(mymap) + geom_point(aes(x = long, y = lat), data = station_locs, 
    alpha = 1, color = "darkred", size = 3)
```


Look at all those stations! Its hard to believe they managed to go out of business!


## Histogram of Number of Bikes per Station

```{r, label = "nbs", fig.cap = "Current dock count", echo = FALSE}
#histogram showing current dock count
ggplot(data = station, aes (x = current_dockcount)) +
geom_bar(fill = "pink", color = "black") +
theme_classic() +
labs( x = "Number of bikes per station", y= "count", title = "Current Dock Count")
```
# hist(station$current_dockcount, breaks = 10, main = "Current Dock Count", xlab = "Number of Bikes Per Station")
# ggplot(station, aes(x = current_dockcount)) + geom_histogram()





```{r}
dock_change <- station %>% group_by(station_id) %>% select(station_id, 
    long, lat, ends_with("dockcount")) %>% mutate(dock_change = current_dockcount - 
    install_dockcount)


#attempt to recreate map plot thing

closeUp <- get_map(location = "Seattle", maptype = "roadmap", zoom = 13)

ggmap(closeUp) + geom_point(aes(x = long, y = lat, size = factor(dock_change), 
                                color = factor(dock_change)), 
                            data = dock_change, alpha = 0.8) + 
  guides(color = guide_legend(title = "change"), size = guide_legend(title = "change")) + 
           scale_size_manual(values = 10:1)




# Change in docking station 
dock_change <- station %>% group_by(station_id) %>% select(station_id, 
long, lat, ends_with("dockcount")) %>% mutate(dock_change = current_dockcount -     install_dockcount)


```

11 stations lost bike docks, 39 docks stayed the same, 8 stations gained docks.


```{r}
# Make the start and stop dates into POSIXct objects
trip_2 <- trip %>% mutate(start_dt = mdy_hm(starttime), stop_dt = mdy_hm(stoptime))

trip_2 <- trip_2 %>% mutate(start_date = paste(month(start_dt), 
     day(start_dt), year(start_dt), sep = "/"))
trip_2$start_date <- mdy(trip_2$start_date)
 
trip_2 <- trip_2 %>% mutate(stop_date = paste(month(stop_dt), 
    day(stop_dt), year(stop_dt), sep = "/"))
trip_2$stop_date <- mdy(trip_2$stop_date)


# Recode the dates
trip_2 %>% 

  group_by(start_date) %>%

  summarize(N = n()) %>%

  ggplot(aes(x = start_date, y = N)) + 

  geom_line() + 

  labs(x = "Date", y = "Number of trips per day") + 

  theme_bw() + geom_smooth()


```


```{r}
#Plotting trips per month (by season, minus summer)
start_date_ym <- trip_2 %>% mutate(ym = paste(year(start_date), 
    month(start_date), sep = "/"), Season = ifelse(ym %in% c("2014/10", "2014/11"), "Fall",
                                            ifelse(ym %in% c("2014/12", "2015/1", "2015/2"), "Winter",
                                            ifelse(ym %in% c("2015/3", "2015/4", "2015/5"), "Spring", "Summer"))))


# ggplot(start_date_ym, aes(x = start_date, y = N))
names(start_date_ym)

# geom_errorbar() needed for this plot


start_date_ym %>%
  group_by(ym, Season) %>%
  summarize(N=n()) %>%
  ggplot(aes(x = ym, y = N, color = Season, fill = Season)) + 
  geom_point() +
  geom_line() + 
  labs( x = "Date")
  



# Convert Trip Duration from Seconds to Minutes
Trip_Duration_Month <- start_date_ym %>% mutate(trip_duration_min = tripduration/60) %>% 
    group_by(ym) %>% select(ym, trip_duration_min) %>% summarise(Avg = mean(trip_duration_min), 
    sd = sd(trip_duration_min)) %>% mutate(se = sd/sqrt(n()))


#Number of Trips by Day of Week
trip_2$wd <- wday(trip_2$start_date, label = TRUE)




tripduration_m <- (trip_2$tripduration)/60

# trip22<-trip_2
# 
# trip22<-cbind(trip22, tripduration_m)


#Trip Duration by Member Type
trip_cost <- trip_2 %>% mutate(cost = ifelse(usertype == "Member" & 
    tripduration_m <= 45, 0, ifelse(usertype == "Member" & tripduration_m > 
    45 & tripduration_m <= 75, 2, ifelse(usertype == "Member" & 
    tripduration_m > 75, (2 + 5 * ((tripduration_m - 75)/30)), 
    ifelse(usertype == "Short-Term Pass Holder" & tripduration_m <= 
        30, 0, ifelse(usertype == "Short-Term Pass Holder" & 
        tripduration_m > 30 & tripduration_m < 60, 2, ifelse(usertype == 
        "Short-Term Pass Holder" & tripduration_m > 60, (2 + 
        5 * ((tripduration_m - 60)/30)), "unknown")))))))

#attempt to make member vs. short term pass holder side by side graph
ggplot(trip_cost, aes(x = usertype)) + facet_grid(cost~.)

```


```{r}
#Members Demographic
trip_2$usertype <- as.factor(trip_2$usertype)
trip_age <- trip_2 %>% mutate(age = year(start_dt) - birthyear)

hist(trip_age$age, main = "Member Age", xlab = "Number of Riders", 
    col = "#56B4E9", breaks = 25)




#Trip routes
# Create a dataframe with only station ID, latitude, and
# longitude
station_coord <- station %>% select(station_id, lat, long)

# Trim our trip dataframe to only include start & stop
# dates/times, and station ID
trip_route <- trip_2 %>% select(trip_id, starts_with("start_"), 
    starts_with("stop_"), from_station_id, to_station_id, tripduration)

# Match by station ID
trip_route$start_lat <- station_coord[match(trip_route$from_station_id, 
    station_coord$station_id), "lat"]

trip_route$start_long <- station_coord[match(trip_route$from_station_id, 
    station_coord$station_id), "long"]

trip_route$stop_lat <- station_coord[match(trip_route$to_station_id, 
    station_coord$station_id), "lat"]

trip_route$stop_long <- station_coord[match(trip_route$to_station_id, 
    station_coord$station_id), "long"]



# Converting trip duration to minutes
trip_route$tripduration <- trip_route$tripduration/60

# Finding actual trip duration
trip_route[1, "tripduration"]


leg_1 %>% group_by(route) %>% summarise(duration = sum(minutes))

trip_route[nrow(trip_route), "tripduration"]



leg_2 %>% group_by(route) %>% summarise(duration = sum(minutes))
```





```{r}
#weather data stuff
str(weather)
# Adjusting the Date Variable
weather$Date <- mdy(weather$Date)

# Adjusting the Events Variable
weather$Events <- as.factor(weather$Events)

#Let’s combine a few of these things that seem to represent the same event.
weather$Events <- gsub("Fog , Rain|Fog-Rain", "Fog-Rain", weather$Events)
weather$Events <- gsub("Rain , Snow|Rain-Snow", "Rain-Snow", 
    weather$Events)
weather$Events <- gsub("Rain , Thunderstorm|Rain-Thunderstorm", 
    "Rain-TS", weather$Events)

weather$Events <- as.factor(weather$Events)


weather$Max_Gust_Speed_MPH <- gsub("-", 0, weather$Max_Gust_Speed_MPH)

weather$Max_Gust_Speed_MPH <- as.numeric(weather$Max_Gust_Speed_MPH)



weather[which(is.na(weather$Mean_Temperature_F)), 1]


weather[490, "Mean_Temperature_F"] <- "50"

weather$Events <- gsub("^$", "Other", weather$Events)
weather$Events <- as.factor(weather$Events)



#weather and trip data combo
# Make a copy of the data frame
trip_3 <- trip_2

# Change column name in trip_3 to match weather dataset
trip_3$Date <- trip_3$start_date

# Left join the trip and weather dataframes by date.
trip_weather <- left_join(trip_3, weather, by = "Date")
```




```{r}
#Analysis

```



```{r}

#our direction

# looking into wind speed and trips
# blah blah blah
```




































































table